\phantomsection
\begin{Acknowledgements}
	First and foremost, I express my profound gratitude to my esteemed supervisors, Prof. Marino Arroyo and Prof. Xavier Trepat, for providing me with the incredible opportunity to embark on this PhD journey and encouraging me to pursue unconventional and fascinating scientific endeavors. Their invaluable guidance, support, and inspiring mentorship have been instrumental in shaping my research and fostering a spirit of curiosity and innovation.
	
	I am deeply grateful to my family for their love, encouragement, and belief in my capabilities. My heartfelt thanks go to my parents, Nilima Chahare and Ramesh Chahare, and sister, Riya Chahare, for their endless support. Particularly, my parents for supporting my decision to move countries for the doctoral studies. Thank you Mummy-Papa. Special thanks to Shubham Pudke for critical mental support and the thesis cover.
	
	My appreciation extends to my closest labmates, the MOLI team, whose camaraderie, collaboration, and intellectual exchange have greatly enriched my experience. I am especially grateful to Tom Golde, Thomas Wilson, and Özge Özguç for their invaluable insights, stimulating discussions, and unwavering assistance during countless hours of experimentation and data analysis. Before Tom joined the project, many things were going wrong. Ideas were present, but the execution and efficiency of experiments improved after he joined. Thomas has been instrumental in trying out new experimental setups and discovering new phenomena. I am truly grateful for having Thomas to discuss new ideas with.
	
	I would like to express my gratitude to Adam Ouzeri. We have been in the same boat, so to speak, throughout the project. He struggled with digital domes while I tackled real ones. Both were challenging. Thank you, Adam. As well as I thank all the lab of Marino Arroyo for developing  computational resources greatly useful in this thesis.
	
	I would like to express my gratitude to all the staff at the IBEC microfabrication facility for providing support with fabrication. Sandra Segura has been instrumental in fabricating new molds for our microfluidic devices. As well as, my gratitude to PRIMO team, a lot of this project would not be possible without them, Dr. Raimon Sunyer and Isabela Fortunato.
	
	The experimental work requires many people, most importantly technicians. Special thanks to Susana Usieto, Natalia Castro, Anghara Mendez, Mònica Purciolas, and Oriol Mañé (my Catalan bro).
	
	The microscope and data management team, which keeps the engines of confocal imaging running, includes Víctor González-Tarragó, Leone Rossetti, Ion Andreu, Ariadna Marin, Ernest Latorre, Raimon Sunyer, Anabelise Le Roux, Amy Beedle, Marija Matejcic, Manuel Gómez, Elena Dalaka, Pau Guillamat, Miquel Bosch Padrós, Nacho Viciano, Tom Golde, and Thomas Wilson.
	
	The cell culture team, responsible for maintaining the conditions that keep my cells alive, consists of Miguel González, Marc Molina, Gerardo Ceada, Jorge Oliver De La Cruz, Sefora Conti, Alice Perucca, Xarxa Quiroga, Anna Labernadie, and Isabela Fortunato. The protocol team and the antibody plasmid team include Aina Albajar, Ignasi Granero Moya, Juanfra Abenza, Mamatha Nijaguna, Ona Baguer, and Meng Wang.
	
	Lastly, the most important team, the pipette team, comprises Srivatsava Viswanadha, Marina Pavlova-Deb, Marilena Naoum, Clément Hallopeau, Kandela Ballerini, Irina Pi i Jaumà and Eva Jou. I would also like to thank Zarina Nauryzgaliyeva for handling the schedules.
	
	Another group of individuals who have enriched my PhD experience are the fellow students around me: Alexandre Garcia-Duran, Joan Triadú, Alexandra Mittens, Gavin McQuarrie, Kandela Ballerini, and Eva Jou.
	
	I would also like to thank Prof. Pere Roca-Cusachs for giving me opportunity to collaborate on number of projects in his lab.
	
	Thesis writing is difficult; but the quality of this document improved with support of MOLI team, Pau Guillamat, Ignasi Granero Moya, Steffen Grosser and Elena Dalaka. Even more difficult to do are the experiments, this is where friendships out of lab mattered a lot. I want to thank Manu Bera, Mohit Bera, Rohit Kosti, Shikhar Nilabh, and Chetan Deshmukh for engaging conversations and great company.
	
	I could not pursue my current line of research without being at the Indian Institute of Science, Bengaluru. I express my gratitude to Prof. Namrata Gundiah for guiding me during my Master's degree and afterwards. I also want to thank Ankur Kulkarni, Somanna Kollimada, Lakshminath Kundanati, Aritra Chatterjee, Neha Paddillaya, and Siddhartha Jaddivada from the lab there.
	
	I would like to extend my heartfelt gratitude to the reviewers, Dr. Raimon Sunyer and Dr. Alejandro Torres-Sánchez, for dedicating their time and providing valuable insights. Additionally, I am deeply appreciative of the examination committee members - Dr. Vikas Trivedi, Dr. José Muñoz, Dr. Magali Suzanne, Dr. David Oriola, and Dr. Raimon Sunyer - for graciously accepting the invitation to participate in my thesis defense.
	
	I extend my gratitude to the Spanish and European governmental grant agencies for providing the funding that made this work possible. I am grateful to the staff at Universitat Politècnica de Catalunya and IBEC for their support. Additionally, I acknowledge the diverse contributions of working people worldwide, from Barcelona's bakers to Pakistani rice farmers, from  Congolese miners to Chinese tech workers, their collective efforts have enabled the exchange of knowledge, goods, and services essential for this work. I am grateful for these unsung heroes of labor, who remind us of the broader context of our academic pursuits and our responsibility to contribute to society and the global community.
	
	Moreover, I acknowledge that this thesis would not have been possible without the support and encouragement of all the people mentioned above, as well as many others who have been part of my journey. I cannot possibly capture everything in these couple of pages. The people with whom I've had fun, engaged in discussions, and formed personal connections will remain with me forever.
	
	Lastly, on a personal note, I extend my deepest appreciation to my partner, Judith Cantó, for her unwavering support, encouragement, and understanding throughout this journey. Her patience, care, and belief in me have been a source of strength and inspiration.
	
	In conclusion, I am forever indebted to the people, and I hope that the knowledge gained through this research would be useful to somebody in the future.
	
	
	
	
	
	
	
	
	
	
	
	
	
	
	
	
	
	
	
	
\end{Acknowledgements}
\cleardoublepage